---
title: "Restricted cubic splines for complex survey data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Restricted cubic splines for complex survey data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6.5,
  fig.height = 4.2,
  dpi = 110
)
```

## Why splines, and why they are awkward here

Real exposure–outcome relationships are rarely straight lines. The usual workarounds all cost
something: categorising a continuous exposure throws away information and power, assuming linearity
misspecifies the model, and polynomials behave badly at the extremes. Restricted cubic splines fit a
smooth curve that is constrained to be linear beyond the outer knots, which keeps the tails stable
while leaving the middle flexible.

Complex surveys such as NHANES make this awkward. Valid analysis needs three design features:

* **sampling weights**, without which point estimates are biased for the population,
* **clustering** and **stratification**, without which standard errors — and therefore every
  confidence interval and *p*-value — are wrong.

`rms`, the natural home of restricted cubic splines in R, does not know about any of them. `survey`
handles all three but has no spline helper. `svyrcs` is the bridge: the spline is fitted *inside* a
`survey` model, so estimates use the weights and the confidence band uses Taylor linearisation on
the design degrees of freedom.

```{r setup, message = FALSE}
library(svyrcs)
library(survey)
library(survival)
```

## The data

The package ships `nhanes_bmi`: adults examined in NHANES 2005–2006 and 2007–2008, linked to the
NCHS 2019 Linked Mortality File. Both cycles are included in full, so every sampling stratum and
PSU is intact.

```{r data}
str(nhanes_bmi, max.level = 1, give.attr = FALSE)
```

Build the design exactly as you would for any other survey analysis. Everything downstream inherits
from it, including the degrees of freedom.

```{r design}
design <- svydesign(
  id = ~psu, strata = ~strata, weights = ~weight,
  nest = TRUE, data = nhanes_bmi
)
degf(design)
```

Thirty-one degrees of freedom, not ten thousand. That single number is the reason a survey analysis
cannot borrow its confidence intervals from an ordinary model fit: the variance is estimated from 62
PSUs in 31 strata, and the intervals have to reflect that.

## A first model

One call. The exposure enters through `rcs()`; everything else is an ordinary model formula.

```{r fit}
fit <- svyrcs(
  Surv(time, event) ~ rcs(bmi, 4) + age + sex + race,
  design = design
)
fit
```

The `Surv()` response selected a survey-weighted Cox model, so the effect measure is a hazard ratio.
Two *p*-values are reported, and they answer different questions:

* **Overall association** — are the spline terms jointly zero? That is, is BMI associated with
  mortality at all?
* **Non-linearity** — are the *non-linear* spline terms jointly zero? A small value here says a
  straight line would not have done, and justifies showing a curve.

Both are design-based *F* tests on `(df1, 31)` degrees of freedom, which is the survey convention
and matches `survey::regTermTest()`. A chi-square version is also stored in `fit$tests`, but it is
anti-conservative with only 31 design degrees of freedom.

## Looking at the curve

```{r plot}
plot(fit, xlab = "Body mass index (kg/m²)")
```

The familiar J: elevated mortality at low BMI, a minimum in the high twenties, and a rise through
the obese range. The dotted vertical line marks the reference value, where the hazard ratio is 1 by
construction.

`summary()` adds what is hard to read off the figure — where the curve turns, and over which
stretches of BMI the confidence band actually excludes 1.

```{r summary}
summary(fit)
```

For numbers to quote in a paper, ask for them directly:

```{r predict}
predict(fit, x = c(18.5, 25, 30, 35, 40))
```

## Choosing the reference

A spline curve is only defined up to a reference: every estimate is a contrast against some exposure
value. `svyrcs` supports the choices seen in the literature, and the choice is purely presentational
— it shifts the curve without changing its shape or either *p*-value.

```{r refs}
fits <- lapply(
  c("median", "min", "max"),
  function(r) svyrcs(Surv(time, event) ~ rcs(bmi, 4) + age + sex,
                     design = design, ref = r)
)
vapply(fits, function(f) f$ref$value, numeric(1))
```

* `"median"` (the default) is the **survey-weighted** population median — not the median of the
  unweighted sample, which would be a property of who happened to be sampled.
* `"min"` anchors at the minimum-risk point, so the whole curve reads as "excess risk relative to
  the healthiest value". This is the most common choice in the obesity literature.
* `"max"`, `"mean"`, `"quantile"` (with `ref_prob`), or any number you like are also accepted.

```{r plot-min}
plot(fits[[2]], xlab = "Body mass index (kg/m²)", title = TRUE)
```

## Other outcomes

The effect measure follows from the model, so nothing about the workflow changes.

```{r families}
# binary outcome: odds ratio
svyrcs(high_chol ~ rcs(bmi, 4) + age + sex, design = design,
       family = quasibinomial())

# continuous outcome: mean difference
svyrcs(tchol ~ rcs(bmi, 4) + age + sex, design = design)
```

Use `quasipoisson()` for rates, which gives rate ratios. The quasi- families are the right choice
under a survey design: the dispersion parameter is not meaningful when the "counts" are weighted, and
`survey` handles the variance itself.

## Does the curve differ by subgroup?

Cross the spline with a grouping variable and `svyrcs()` fits **one** model, returning a curve per
group:

```{r subgroup}
by_sex <- svyrcs(Surv(time, event) ~ rcs(bmi, 4) * sex + age, design = design)
by_sex
```

```{r subgroup-plot}
plot(by_sex, xlab = "Body mass index (kg/m²)")
```

Two questions get separate answers. **Interaction** asks whether the association differs between
groups at all. **Shape interaction** drops the linear interaction term, so it asks the narrower
question: does the *curvature* differ, or is one curve simply a shifted copy of the other?

Here neither is close to significant, and that is worth dwelling on. The two curves in the figure
look different — the minimum sits a couple of BMI units higher in women, and the male curve climbs
sooner — but with 31 design degrees of freedom that separation is well inside the noise. Reporting
"the association was stronger in men" from this fit would not be supported. Subgroup curves are easy
to over-read, which is exactly why the interaction *p*-value is printed next to them.

`facet = TRUE` gives each group its own panel when the bands overlap too much to read:

```{r subgroup-facet, fig.height = 3.4}
plot(by_sex, facet = TRUE, xlab = "Body mass index (kg/m²)")
```

Some details worth knowing:

* One model, not several. Covariate effects — `age` here — are shared across groups. Fitting each
  subgroup separately would estimate them independently and could not produce an interaction test at
  all.
* All groups share one reference value, so the curves are comparable. With `ref = "min"` the
  reference is the minimum-risk point of the *reference level*, and the output names it.
* The standard error for a non-reference group accounts for the covariance between the main and
  interaction coefficients. Adding the two variances separately, which is the obvious shortcut,
  gives the wrong interval.
* The modifier must be a factor, character or logical. Continuous effect modification is not
  supported.
* `predict(fit, x = ..., group = "Female")` gives numbers for one group.

## Missing data

The shipped data has missing cholesterol for 649 people. Everything above quietly dropped them.
Multiple imputation is the usual alternative, and `svyrcs` consumes the result: build the
imputations however you like, wrap them in `mitools::imputationList()`, and pass the resulting
design.

```{r mi, eval = requireNamespace("mice", quietly = TRUE) && requireNamespace("mitools", quietly = TRUE), message = FALSE}
library(mice)
library(mitools)

vars <- c("age", "sex", "race", "bmi", "tchol", "hdl", "time", "event")
imp <- mice(nhanes_bmi[vars], m = 5, method = "pmm", maxit = 3, printFlag = FALSE, seed = 1)

completed <- lapply(1:5, function(i) {
  cbind(complete(imp, i), nhanes_bmi[c("psu", "strata", "weight")])
})

mi_design <- svydesign(
  id = ~psu, strata = ~strata, weights = ~weight,
  nest = TRUE, data = imputationList(completed)
)

fit_mi <- svyrcs(Surv(time, event) ~ rcs(bmi, 4) + age + sex + tchol, design = mi_design)
fit_mi
```

Two things are worth understanding here.

**Knots and the reference are fixed before any model is fitted**, as the average of the
survey-weighted quantiles across imputations. This is not a convenience: if each imputation chose its
own knots, the five coefficient vectors would be estimates of five different parameterisations and
Rubin's rules would not apply to them.

**The degrees of freedom are not `mitools`'.** `MIcombine()` reports `(m-1)(1+1/r)^2`, which has no
upper bound — on this design, which has 31 degrees of freedom, it returns numbers in the thousands
and sometimes `Inf`. Feeding that to a *t* quantile would give confidence intervals far too narrow,
the same error as using a normal quantile but worse. `svyrcs` applies the Barnard-Rubin correction,
which is bounded by the complete-data degrees of freedom:

```{r mi-df, eval = requireNamespace("mice", quietly = TRUE) && requireNamespace("mitools", quietly = TRUE)}
head(predict(fit_mi, x = c(18, 22, 27, 35, 45)))
```

The `df` column sits at or below 31 and drops where the fraction of missing information (`fmi`) is
highest. When nothing in the model is imputed, `fmi` is zero, `df` is exactly 31, and the result is
identical to the complete-data fit.

The joint tests are treated the same way. Their denominator degrees of freedom are adjusted rather
than taken from the usual multiple-imputation rule, which is derived from the imputation structure
alone and ignores the survey design.

## Knots

Four knots at Harrell's recommended quantiles is the default, and is a reasonable choice for most
sample sizes. More knots buy flexibility at the cost of stability.

```{r knots}
fit$knots
svyrcs(Surv(time, event) ~ rcs(bmi, 5) + age, design = design)$knots
svyrcs(Surv(time, event) ~ rcs(bmi, c(20, 25, 30, 40)) + age, design = design)$knots
```

By default the quantiles are **survey-weighted**, so knot placement reflects the population
distribution. `weighted_knots = FALSE` reproduces the unweighted placement you would get from a
naive script.

Whatever you choose, the knots are stored on the fitted model, so predictions on new data can never
be made against silently re-derived knots:

```{r predvars}
fit$knots
attr(terms(fit$model), "predvars")[[2]]
```

## Working from a model you fitted yourself

If you need a specification `svyrcs()` does not build — a stratified Cox model, an unusual family,
extra arguments — fit it yourself and hand it over:

```{r curve-direct}
m <- svycoxph(Surv(time, event) ~ rcs(bmi, 4) + age + strata(sex), design = design)
head(svyrcs_curve(m, "bmi", ref = "min", n = 5))
```

## How the estimates are computed

Worth knowing, because it differs from the hand-rolled scripts that circulate for this task.

The curve is a contrast in the linear predictor between exposure `x` and the reference `x0`. Because
the linear predictor is additive and every covariate takes the *same* value at both points, all
covariate contributions cancel exactly. Only the spline columns remain:

```
dB     = B(x) − B(x₀)
effect = dB β,   var(effect) = dB V dBᵀ
```

with `β` and `V` the spline coefficients and their design-based covariance. This is exact, not an
approximation — it reproduces `predict(type = "lp")` differences to about 1e-15 — and it means there
is no need to build a prediction grid with covariates fixed at reference values. That matters
practically: such grids break when a covariate is a factor with a level absent from the grid, and
they invite the error of comparing across different covariate profiles.

Confidence intervals then use a *t* quantile on `degf(design)` rather than a normal quantile. With
31 degrees of freedom that is a 4% wider interval, which is not nothing.

## Limitations

* Negative binomial outcomes are not supported: `survey` has no such family. Use `quasipoisson()`,
  whose variance is estimated from the design anyway.
* One spline exposure per model. Fit separate models, or use `svyrcs_curve()` on your own fit.
* Effect modification by a *continuous* variable, and three-way interactions, are not supported.
  Group modifiers are (see above).
* Analyses of a subsample with its own weight (for instance the NHANES fasting subsample) need that
  subsample's weight in `svydesign()`; the shipped data carries only the MEC exam weight.

## References

1. Harrell FE Jr. *Regression Modeling Strategies*. 2nd ed. Springer; 2015.
   \doi{10.1007/978-3-319-19425-7}
2. Lumley T. *Complex Surveys: A Guide to Analysis Using R*. Wiley; 2010.
   \doi{10.1002/9780470580066}
3. Binder DA. On the variances of asymptotically normal estimators from complex surveys.
   *International Statistical Review*. 1983;51(3):279–292. \doi{10.2307/1402588}
4. CDC. [NHANES tutorials](https://wwwn.cdc.gov/nchs/nhanes/tutorials/default.aspx).
